\documentclass[article]{jss}

\usepackage{amsmath} % For mathematical equations
\usepackage{amsfonts} % For mathematical fonts
\usepackage{amssymb} % For mathematical symbols
\usepackage{graphicx} % For including graphics
\usepackage{hyperref} % For hyperlinks
\usepackage{url} % For better URL formatting
\usepackage{enumitem} % For custom list environments
\usepackage{float} % For precise figure placement
\usepackage{booktabs} % For professional tables
\usepackage{listings} % For code listings (or use minted package if you have a LaTeX distribution with Pygments)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Nathan Green \\Department of Statistical Science,\\ UCL, London, UK \And 
        Antonio Remiro-Azocar \\Novo Nordisk}
\title{Population-Adjusted Indirect Treatment Comparison with the \pkg{outstandR} Package in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Nathan Green, Antonio Remiro-Azocar} %% comma-separated
\Plaintitle{Indirect Treatment Comparison with the outstandR Package in R} %% without formatting
\Shorttitle{\pkg{outstandR}: Indirect Treatment Comparison} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Indirect treatment comparisons (ITCs) are essential in Health Technology Assessment (HTA) when head-to-head clinical trials are absent. A common challenge arises when attempting to compare a treatment with available individual patient data (IPD) against a competitor with only reported aggregate-level data (ALD), particularly when trial populations differ in effect modifiers. While methods such as Matching-Adjusted Indirect Comparison (MAIC) and Simulated Treatment Comparison (STC) exist to adjust for these cross-trial differences, software implementations have often been fragmented or limited in scope. This article introduces \pkg{outstandR}, an \proglang{R} package designed to provide a comprehensive and unified framework for population-adjusted indirect comparison (PAIC). Beyond standard weighting and regression approaches, \pkg{outstandR} implements advanced G-computation methods—within both maximum likelihood and Bayesian frameworks, and Multiple Imputation Marginalization (MIM) to address non-collapsibility and missing data. By streamlining the workflow of covariate simulation, model standardization, and contrast estimation, \pkg{outstandR} enables robust and compatible evidence synthesis in complex decision-making scenarios.
}
\Keywords{G-computation, simulated treatment comparison, matching-adjusted indirect comparison, Bayesian, multiple imputation, \proglang{R}}
\Plainkeywords{g-computation, simulated treatment comparison, matching-adjusted indirect comparison, bayesian, multiple imputation, R} %% without formatting

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Nathan Green\\
  Department of Statistical Science\\
  UCL\\
  E-mail: \email{n.green@ucl.ac.uk}\\
  URL: \url{https://n8thangreen.github.io/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Define a minimalist style that only sets the font to ttfamily
\lstdefinestyle{textttstyle}{
    basicstyle=\ttfamily, % Set the font to typewriter
    breaklines=true,       % Allow lines to break
    columns=fullflexible   % Better for line wrapping
}

\begin{document}

\section{Introduction}
\label{sec:introduction}

Population adjustment methods are increasingly used to compare marginal treatment effects when there are cross-trial differences in effect modifiers and limited patient-level data \cite{RemiroAzocar2022b}. This presents a significant challenge in evidence synthesis, particularly in health technology assessment. Existing methods often face limitations such as sensitivity to poor covariate overlap or inability to extrapolate beyond observed covariate spaces.

The \pkg{outstandR} package addresses these challenges by offering a robust framework for outcome regression standardisation, focusing on G-estimation. It enables researchers to perform model-based standardization with two additional crucial steps: covariate simulation (to overcome limited subject-level data for one of the studies) and indirect comparison across studies \cite{RemiroAzocar2022a}.

Currently, practitioners seeking to perform model-based standardization must often rely on ad-hoc scripts or fragmented software tools that lack a unified interface. \pkg{outstandR} targets this inefficiency by streamlining the entire workflow — from covariate simulation to contrast estimation — within a single, cohesive, extensible package.

% outline
The remainder of this paper is organized as follows. Section~\ref{} 2 establishes the methodological framework, defining the general problem of indirect treatment comparisons and detailing the statistical mechanics of population adjustment methods, including Matching-Adjusted Indirect Comparison (MAIC), Simulated Treatment Comparison (STC), and advanced G-computation approaches. Section~\ref{} 3 describes the software architecture of outstandR, illustrating how the package leverages an S3 object-oriented design to separate statistical strategies from computational execution and streamline the workflow. Section~\ref{} 4 provides a practical demonstration of the package, guiding the user through illustrative examples using synthetic datasets for both binary and continuous outcomes. Finally, Section~\ref{} 5 concludes with a discussion of the package's capabilities and potential areas for future development.

\section{Population-Adjusted Indirect Comparison (PAIC)}
\label{sec:methodology}
Consider a scenario with two trials: an $AC$ trial with individual patient data (IPD) comparing treatments $A$ and $C$, and a $BC$ trial with only published aggregate-level data (ALD) comparing treatments $B$ and $C$. The objective is to estimate a comparison of the effects of treatments $A$ and $B$ on a suitable scale in some target population $P$, denoted by the parameter $d_{AB(P)}$ \cite{RemiroAzocar2022a}. We will focus on population adjustment methods to make ``anchored'' indirect comparisons, where the evidence is connected by a common comparator, in our case $C$, and ``unanchored'' indirect comparisons, where the evidence is disconnected due to a lack of a common comparator or single-arm studies.

\subsection{General problem formulation}
\label{ssec:problem_formulation}
Let $\mu_{A(AC)}$ and $\mu_{C(AC)}$ represent the expected outcomes for treatments $A$ and $C$ within the $AC$ trial population. We also have estimators $\bar{Y}_{A(AC)}$ and $\bar{Y}_{C(AC)}$ of the expected outcomes. There is also an equivalent set of parameters for the $B$ trial. 

We define estimators $\Delta_{AC(AC)}$ and $\Delta_{BC(BC)}$ of the trial-level relative treatment effects for a given scale. Assuming a {\it difference} scale (e.g., log-odds ratio, risk difference), the estimate of relative treatment effect between treatments $A$ and $C$
\begin{equation}
\label{eqn:relative_treatment_effect}
\Delta_{AC(AC)} = g(\bar{Y}_{C{(AC)}}) - g(\bar{Y}_{A{(AC)}}),
\end{equation}
and similarly for $\Delta_{BC(BC)}$. If we assume that there is no difference in effect modifiers between trials, then the estimator of the relative treatment effect $d_{AB(BC)}$ is
\[
\Delta_{AB(BC)} = \Delta_{BC(BC)} - \Delta_{AC(AC)}.
\]
However, when distributions of the effect modifiers are different between trial populations, the relative treatment effect estimated from each trial cannot simply be combined as above. This is the purpose of population-adjustment in ITC. Section~\ref{sec:paic-methods} will detail the PAIC methods available in \pkg{outstandR}. Before this, we present the specific formulae for `standard' treatment comparison for different data types and scales.

\subsection{Relative treatment effects}
This section details the equations used to calculate the mean and variance for the treatment effect estimator $\Delta_{\cdot \cdot (P)}$ in Equation~(\ref{eqn:relative_treatment_effect}) specific for various data types and outcome scales, as implemented in \pkg{outstandR}. These calculations are fundamental for making indirect treatment comparisons.

For each treatment $t$, define the following:
\begin{itemize}
    \item $p_t$: Observed proportion (for binary outcomes).
    \item $\bar{y}_t$: Observed mean (for continuous or count outcomes).
    \item $s_t$: Observed standard deviation (for continuous outcomes).
    \item $N_t$: Number of observations.
\end{itemize}

\subsubsection{Mean treatment effects}
Using aggregate data summary statistics we can calculate different types of relative effects for binary, continuous and count outcomes.

\paragraph{Binary Outcomes}
For binary outcomes, let $p_C$ be the proportion of events in the comparator arm and $p_B$ be the proportion of events in the reference arm.
\begin{align*}
    \text{Log-Odds Ratio:} \quad & \text{logit}(p_C) - \text{logit}(p_B) \\
    \text{Risk Difference:} \quad & p_C - p_B \\
    \text{Probit Scale Difference ($\Delta z$):} \quad & \text{probit}(p_C) - \text{probit}(p_B) \\
    \text{Log Relative Risk (Rare Events - cloglog):} \quad & \ln(-\ln(1 - p_C)) - \ln(-\ln(1 - p_B)) \\
    \text{Log Relative Risk (Log Link):} \quad & \ln(p_C) - \ln(p_B)
\end{align*}

\paragraph{Continuous outcomes}
For continuous outcomes, let $\bar{y}_C$ be the mean in the comparator arm and $\bar{y}_B$ be the mean in the reference arm.
\begin{align*}
    \text{Mean Difference:} \quad & \bar{y}_C - \bar{y}_B
\end{align*}
For other scales on continuous data, the approach may involve transforming the means directly, though typical continuous outcome analyses focus on mean differences.

\paragraph{Count outcomes}
For count outcomes, let $\bar{y}_C$ be the mean count or rate in the comparator arm and $\bar{y}_B$ be the mean count/rate in the reference arm.
\begin{align*}
    \text{Log Relative Risk (Log Link):} \quad & \ln(\bar{y}_C) - \ln(\bar{y}_B) \\
    \text{Rate Difference:} \quad & \bar{y}_C - \bar{y}_B \\
    \text{Delta $z$ (Signal-to-noise):} \quad & \frac{\bar{y}_C}{\sqrt{\bar{y}_C}} - \frac{\bar{y}_B}{\sqrt{\bar{y}_B}} = \sqrt{\bar{y}_C} - \sqrt{\bar{y}_B}
\end{align*}

\subsubsection{Variance of treatment effects}
The variance of the treatment effect for each arm is calculated based on the distribution family and the chosen effect scale. Let $\text{Var}(\Delta_t)$ denote the variance contribution from treatment $t$. The variance of the treatment effect for a direct comparison (e.g., $C$ vs $B$) would then be $\text{Var}(\Delta_{C}) + \text{Var}(\Delta_{B})$.

\paragraph{Binary outcomes}
Let $y_t$ be the number of events and $N_t$ be the total number of subjects for treatment $t$. Then $\text{Var}(\Delta_t)$ is
\begin{align*}
    \text{Log-Odds Ratio:} \quad & \frac{1}{y_t} + \frac{1}{N_t - y_t} \\
    \text{Log Relative Risk:} \quad & \frac{1}{y_t} - \frac{1}{N_t} \\
    \text{Risk Difference:} \quad & \frac{p_t(1 - p_t)}{N_t} \\
    \text{Delta $z$:} \quad & \frac{1}{y_t} + \frac{1}{N_t - y_t} \\
    \text{Log Relative Risk (Rare Events):} \quad & \frac{1}{y_t} - \frac{1}{N_t}
\end{align*}

\paragraph{Continuous outcomes}
Let $\bar{y}_t$ be the mean, $s_t$ be the standard deviation, and $N_t$ be the number of subjects for treatment arm $t$.
\begin{align*}
    \text{Mean Difference:} \quad & \frac{s_t^2}{N_t}
\end{align*}
For log-odds and log-relative risk on continuous data, the \proglang{R} code implies a transformation of the mean before calculating the variance. For ``log\_odds'', it uses $\frac{\pi^2}{3 N_t}$, which is the variance of a logit-transformed mean when the underlying data is assumed to be normal and mapped to (0,1) scale, and then transformed to logit (often used for proportions derived from continuous data or for scores that can be thought of as probabilities).
% For ``log\_relative\_risk'', the \proglang{R} code simply returns $\log(\bar{y})$, which appears to be a placeholder or an indication that the variance is derived differently or approximated in specific contexts.
We assume here that for continuous data, the primary interest is the mean difference and its variance.

\paragraph{Count outcomes}
Let $\bar{y}_t$ be the mean count or rate and $N_t$ be the number of subjects or exposure for treatment arm $t$.
\begin{align*}
    \text{Log Relative Risk:} \quad & \frac{1}{N_t \bar{y}_t} \\
    \text{Rate Difference:} \quad & \frac{\bar{y}_t}{N_t} \\
    \text{Delta $z$ (Signal-to-noise):} \quad & \frac{1}{4 N_t \bar{y}_t}
\end{align*}

\subsection{Population adjustment methods}
\label{sec:paic-methods}
The \pkg{outstandR} package allows the implementation of a range of methods for adjusting the IPD to the target population, commonly the ALD.

We make a clear and necessary distinction between prognostic factors (PF) and effect modifiers (EM). Prognostic factors are covariates that affect the outcome whereas effect modifiers are covariates that alter the effect of treatment as measured on a given scale. Effect modifiers are not necessarily also prognostic variables, and may be specific to each treatment. Note that effect modifier status on one scale does not necessarily imply effect modifier status on another scale. We also assume internal validity of the studies included in the analysis. That is, the studies provide unbiased estimates of treatment effects in their respective sample populations.

\subsubsection{Matching-Adjusted Indirect Comparison (MAIC)}
Matching-Adjusted Indirect Comparison (MAIC) is based on propensity score weighting, reweights IPD to match aggregate characteristics of the comparator trial \cite{RemiroAzocar2022a}. The core idea of MAIC is to re-weight the patients in the IPD $AC$ trial such that their aggregated baseline characteristics match those of the ALD $BC$ trial. This re-weighting process is essentially an optimization problem.

Let $X_i$ be the vector of baseline characteristics for the $i$-th patient in the IPD $AC$ trial, for $i = 1, \dots, n_A$, where $n_A$ is the number of patients in $AC$ trial. Let $\bar{X}_B$ be the vector of mean baseline characteristics reported for the ALD $BC$ trial.

The objective is to find a set of weights, $w_i$, for each patient $i$ in Trial A, such that the weighted average of their baseline characteristics matches the aggregate baseline characteristics of $BC$ trial. This can be formulated as minimizing a certain objective function, typically related to the sum of weights or the effective sample size, subject to matching constraints.

A common approach involves a logistic regression-based re-weighting. The weights $w_i$ are often derived from a propensity score-like model, where the parameters of this model are optimized. Specifically, the weights are defined as:

$$ w_i = \exp(\beta^T X_i) $$
where $\beta$ is a vector of coefficients to be estimated. The optimization problem then aims to find $\beta$ such that the weighted means of the covariates in the IPD dataset match the target means from the ALD dataset. This is equivalent to solving the following set of moment-matching equations:

$$ \sum_{i=1}^{n_A} w_i X_i = \sum_{i=1}^{n_A} \exp(\beta^T X_i) X_i = n_A \bar{X}_B $$
This system of equations can be rewritten by centering the covariates around the target means $\bar{X}_B$:

$$ \sum_{i=1}^{n_A} \exp(\beta^T (X_i - \bar{X}_B)) (X_i - \bar{X}_B) = 0 $$
This is the first derivative with respect to $\beta$ of the following convex objective function, which is typically minimized:

$$ \min_{\beta \in \mathbb{R}^p} \sum_{i=1}^{n_A} \exp(\beta^T (X_i - \bar{X}_B)) $$
where $p$ is the number of baseline characteristics being matched. The minimization of this convex function ensures a unique global minimum for $\beta$. Once $\beta$ is found, the weights $w_i$ are calculated, and these weights are then used in a weighted analysis (e.g., weighted logistic regression or Cox regression) of the IPD from the $AC$ trial to estimate the treatment effect in a population that is adjusted to resemble the $BC$ trial.

The resulting effect size from the weighted IPD in the $AC$ trial can then be indirectly compared to the treatment effect observed in Trial B (often compared to its common comparator in an ``anchored" MAIC, or as a direct comparison in ``unanchored" MAIC). The optimization process ensures that the populations being compared are as similar as possible based on the available covariates, thereby reducing potential bias due to differences in prognostic factors or effect modifiers.

\subsubsection{Simulated Treatment Comparison (STC)}
Simulated Treatment Comparison (STC) relies on outcome regression models fitted to IPD, conditioning on covariates, then applying estimates to ALD \cite{RemiroAzocar2022a}. STC involves using the IPD to build a model that predicts patient outcomes based on their baseline characteristics, and then using this model to estimate what the outcomes of the IPD trial's treatment would have been if it were applied to the population of the ALD $BC$ trial.

\paragraph{Model Fitting (Optimization Step)}
Let $Y_i$ be the outcome for patient $i$ in the IPD $AC$ trial, and $X_i$ be the vector of baseline characteristics for patient $i$. An outcome regression model is fitted using the IPD $AC$ trial. This model relates the outcome to the baseline covariates. For a continuous outcome, a linear regression might be used; for a binary outcome, logistic regression; and for time-to-event data, a Cox proportional hazards model or parametric survival model.

The general form of this outcome model can be expressed as:
$$ g(\mathbb{E}[Y_i | X_i]) = \alpha + \beta^T X_i $$
where
\begin{itemize}
    \item $g(\cdot)$ is a link function (e.g., identity for linear regression, logit for logistic regression, log for Cox regression hazard ratio).
    \item $\mathbb{E}[Y_i | X_i]$ is the expected outcome for patient $i$ given their baseline characteristics $X_i$.
    \item $\alpha$ is the intercept.
    \item $\beta$ is a vector of regression coefficients for the baseline characteristics $X_i$.
\end{itemize}

The ``optimization" in this step refers to the standard process of estimating the model parameters ($\alpha$ and $\beta$) by maximizing the likelihood function (or minimizing a sum of squares, depending on the model).

\paragraph{Prediction (Simulation Step)}
Once the outcome model is fitted using data from Trial A (IPD), the estimated coefficients ($\hat{\alpha}, \hat{\beta}$) are used to predict the expected outcome of Treatment A for the population of Trial B. Since only aggregate data (typically means and standard deviations) are available for Trial B's population, these aggregate values ($\bar{X}_B$) are substituted into the fitted model to estimate the mean outcome.

The predicted mean outcome of treatment $A$ in the population of the $BC$ trial, denoted as $\hat{\mathbb{E}}[Y_A | \text{Pop}_B]$, is calculated as:
$$
\hat{\mathbb{E}}[Y_A | \text{Pop}_B] = g^{-1}(\hat{\alpha} + \hat{\beta}^T \bar{X}_B)
$$
where $g^{-1}(\cdot)$ is the inverse of the link function.

\paragraph{Limitations and modifications}
A significant limitation of standard STC arises when the link function $g(\cdot)$ is non-linear (e.g., logit or log link). In such cases, simply substituting the mean covariate values ($\bar{X}_B$) into the fitted model can lead to biased estimates of the marginal treatment effect in the target population (this is known as ``aggregation bias'' or ``non-collapsibility'').

To address this, more advanced STC methods, particularly those incorporating actual simulation, are being developed. These methods aim to correctly estimate the marginal outcome in the target population, often by simulating individual patient profiles for the ALD trial's population based on their reported aggregate statistics and assumed covariate distributions. This more sophisticated approach is truly simulation-based and involves drawing from estimated joint covariate distributions and leads into the following methods based on G-computation.

\subsubsection{Parametric G-computation with Maximum Likelihood}
G-computation, also known as the G-formula, fits an outcome model to IPD using maximum likelihood, then predicts outcomes in the comparator population \cite{RemiroAzocar2022a}. It is a method for estimating causal effects by standardizing outcomes across different treatment (or exposure) regimens. It directly models the conditional expectation of the outcome given exposure and covariates. When implemented using Maximum Likelihood Estimation (MLE), the process involves fitting a series of regression models for the outcome and, potentially, for time-varying confounders, and then using these models to predict outcomes under hypothetical exposure scenarios.

The fundamental idea of G-computation is to emulate a randomized trial by predicting outcomes under different treatment assignments for each individual in the observed (or a representative) population, and then averaging these predicted outcomes.

The target parameter for G-computation is often the marginal mean outcome under a specific treatment regimen $A^*$, defined as:
$$
\mathbb{E}[Y^{A^*}] = \sum_{x} \mathbb{E}[Y | A^*=a, X=x] P(X=x)
$$
For continuous covariates, the sum becomes an integral. In essence, it calculates the average outcome if everyone in the population followed the treatment regimen $A^*$.

\paragraph{Maximum Likelihood Implementation Steps}
The MLE implementation of G-computation typically involves the following steps:

First, specify and fit models for outcome and confounders. This is the ``optimization" step where parameters are estimated using MLE. A set of regression models are specified for the outcome and any time-varying confounders.

\begin{enumerate}
    \item Outcome Model: A model for the outcome $Y$ given the treatment $A$ (or sequence of treatments $A_t$) and relevant covariates $L$ (baseline and/or time-varying confounders).
    $$
    \mathbb{E}[Y | A, L] = h(\theta_Y^T Z_Y)
    $$
    where $h(\cdot)$ is the inverse of the link function, $Z_Y$ is a vector of covariates and treatment terms, and $\theta_Y$ are the parameters to be estimated. For example, for a continuous outcome $Y$:
    $$ Y \sim N(\beta_0 + \beta_A A + \beta_L L + \dots, \sigma^2) $$
    The parameters $(\beta_0, \beta_A, \beta_L, \dots, \sigma^2)$ are estimated by maximizing the likelihood of the observed data.
    \item Confounder Models (for time-varying confounders): If there are time-varying confounders $L_t$ that are themselves affected by prior treatment and affect future outcomes, separate models are fitted for each $L_t$ given prior covariates and treatment history.
    $$
    \mathbb{E}[L_t | A_{0 \dots t-1}, L_{0 \dots t-1}] = f(\theta_{L_t}^T Z_{L_t})
    $$
    where $f(\cdot)$ is a link function, $Z_{L_t}$ includes relevant past treatment and covariate history, and $\theta_{L_t}$ are the parameters estimated by maximizing the likelihood for each confounder model.
\end{enumerate}
The estimation of these parameters ($\theta_Y, \theta_{L_t}$) constitutes the maximum likelihood optimization. For each model, we seek the parameters that maximize the likelihood function $\mathcal{L}(\theta | \text{Data})$, or equivalently, minimize the negative log-likelihood.

\paragraph{Simulation/Prediction Step}
Once all necessary models are fitted, the estimated parameters are used to predict outcomes under a hypothesized treatment. This can be done via:

\begin{enumerate}
    \item Monte Carlo Simulation (typically for complex scenarios):
    \begin{itemize}
        \item For each individual in the original dataset (or a large simulated population representative of it):
        \item Set their treatment to the hypothesized regimen $A^*$.
        \item Use the fitted confounder models to simulate trajectories of time-varying confounders under this regimen.
        \item Use the fitted outcome model to predict the outcome for each individual under the simulated covariate and treatment path.
        \item Average the predicted outcomes across all individuals to obtain the marginal mean outcome for the treatment $A^*$.
    \end{itemize}

    \item Direct Standardization (for simpler cases without time-varying confounding):
    \begin{itemize}
        \item For each observed individual $i$, predict their outcome under the hypothesized treatment $A^*$:
        $$ \hat{Y}_i(A^*) = h(\hat{\theta}_Y^T Z_Y(A^*, L_i)) $$
        where $Z_Y(A^*, L_i)$ incorporates the hypothesized treatment $A^*$ for individual $i$ with their observed covariates $L_i$.
        \item The estimated marginal mean outcome is then the average of these predictions:
        $$
        \hat{\mathbb{E}}[Y^{A^*}] = \frac{1}{N} \sum_{i=1}^N \hat{Y}_i(A^*) $$
    \end{itemize}
\end{enumerate}

\subsubsection{Parametric G-computation with Bayesian Inference}
G-computation (G-formula) can be implemented using Bayesian inference, Similar to the maximum likelihood version, but within a Bayesian framework, allowing for coherent propagation of uncertainty \cite{RemiroAzocar2022a}.
This offers several advantages over the frequentist Maximum Likelihood Estimation (MLE) approach, particularly in quantifying uncertainty and incorporating prior knowledge. Instead of finding point estimates for model parameters that maximize a likelihood function, the Bayesian approach involves estimating the full posterior distribution of these parameters. This allows for direct probability statements about the parameters and the resulting causal effects.

The goal remains the same: to estimate the marginal mean outcome under a specific treatment regimen $A^*$, defined as $\mathbb{E}[Y^{A^*}]$. The core idea is to average predicted outcomes under counterfactual treatment scenarios.

The Bayesian implementation of G-computation typically follows these steps:

\paragraph{Step 1: Specify and Fit Models with Bayesian Inference}
First, specify and fit models with Bayesian inference. This is where the ``optimization" (in a Bayesian sense, parameter estimation via sampling) occurs. For each model (outcome and time-varying confounders), we define a likelihood function and specify prior distributions for all parameters.

\begin{enumerate}
    \item Model Specification: Define the same types of regression models as in the frequentist approach for the outcome
        $$
        Y_i \sim g^{-1}(\alpha + \beta^T X_i)
        $$
        where $g^{-1}(\cdot)$ is the inverse of the link function, $X_i$ includes treatment and covariate terms, and $\alpha, \beta$ are the parameters.

    \item Prior Specification: Assign prior distributions to all model parameters (e.g., $\alpha, \beta, \gamma$, and any dispersion parameters like $\sigma^2$). These priors reflect existing knowledge or assumptions about the parameters before observing the data. For example:
    $$ \alpha \sim N(0, \sigma_{\alpha}^2) $$
    $$ \beta_k \sim N(0, \sigma_{\beta}^2) \quad \forall k $$
    $$ \sigma^2 \sim \text{Inv-Gamma}(a, b) $$

    \item Posterior Sampling (Optimization/Inference):
    Using computational methods like Markov Chain Monte Carlo (MCMC) (e.g., Gibbs sampling, Metropolis-Hastings, Hamiltonian Monte Carlo), samples are drawn from the joint posterior distribution of all model parameters.
    This process, while not a direct maximization, effectively ``optimizes" in the sense of finding the region of highest posterior probability, and importantly, characterizes the uncertainty around these parameters. The result is a set of $M$ posterior samples for each parameter, e.g., $(\alpha^{(m)}, \beta^{(m)}, \gamma^{(m)})$, for $m=1, \dots, M$.
\end{enumerate}

\paragraph{Step 2: Prediction/Simulation Step (Posterior Predictive Simulation)}
For each set of sampled parameters from the posterior distribution ($m=1, \dots, M$):
\begin{enumerate}
    \item Generate Counterfactual Data:
    \begin{itemize}
        \item For each individual $i$ in the observed population (or a large simulated population representative of it), set their treatment to the hypothesized regimen $A^*$.
        \item Using the $m$-th sample of parameters for the confounder models ($\gamma^{(m)}$), simulate trajectories of time-varying confounders under this regimen. This involves drawing from the conditional predictive distributions of $L_t$ for each individual, sequentially.
        \item Using the $m$-th sample of parameters for the outcome model ($\alpha^{(m)}, \beta^{(m)}$), predict the outcome for each individual under the simulated covariate and treatment path. This also involves drawing from the conditional predictive distribution of $Y$.
    \end{itemize}

    \item Calculate Marginal Mean for this Sample:
    Average the predicted outcomes across all individuals to obtain the marginal mean outcome for the regimen $A^*$ for this $m$-th posterior sample:
    $$
    \hat{\mathbb{E}}[Y^{A^*}]^{(m)} = \frac{1}{N} \sum_{i=1}^N \hat{Y}_i(A^*)^{(m)}
    $$
\end{enumerate}
This results in $M$ samples of the marginal mean outcome $\hat{\mathbb{E}}[Y^{A^*}]^{(m)}$, which together form the posterior distribution of the marginal mean outcome under regimen $A^*$.

\paragraph{Step 3: Causal Effect Estimation}
To estimate the causal effect of two different regimens, $A_1^*$ and $A_0^*$, the above process is repeated to obtain $M$ posterior samples for each: $\hat{\mathbb{E}}[Y^{A_1^*}]^{(m)}$ and $\hat{\mathbb{E}}[Y^{A_0^*}]^{(m)}$.

The posterior distribution of the causal effect is then directly obtained by computing the difference or ratio for each posterior sample:
$$
\Delta^{(m)} = \hat{\mathbb{E}}[Y^{A_1^*}]^{(m)} - \hat{\mathbb{E}}[Y^{A_0^*}]^{(m)} \quad \text{ (for differences)}
$$
$$
\Delta^{(m)} = \frac{\hat{\mathbb{E}}[Y^{A_1^*}]^{(m)}}{\hat{\mathbb{E}}[Y^{A_0^*}]^{(m)}} \quad \text{ (for ratios)}
$$
From this posterior distribution of the causal effect, one can derive point estimates (e.g., median), credible intervals (e.g., 2.5th and 97.5th percentiles), and posterior probabilities (e.g., probability that the effect is greater than zero).

Key advantages of Bayesian G-computation are full uncertainty quantification: Provides direct posterior distributions for all parameters and causal effects, allowing for more comprehensive uncertainty assessment. Incorporation of prior knowledge by allowing for the inclusion of expert knowledge or evidence from previous studies through informative prior distributions. Handling complex models since MCMC methods are well-suited for fitting complex hierarchical or non-linear models often encountered in causal inference. Direct probabilistic statements like ``there is a 95\% probability that the causal effect is between X and Y."


\subsubsection{Multiple Imputation Marginalization Method based on Parametric G-computation}
Multiple Imputation Marginalization (MIM) aims to recover a compatible marginal treatment effect by integrating or averaging the conditional estimate over the relevant population, particularly useful for generalized linear models or Cox models \cite{RemiroAzocar2022b}.
This is a robust method for estimating marginal causal effects, particularly within the framework of parametric G-computation, while simultaneously addressing missing data in the covariates or outcome. Developed by Remiro-Azócar et al., MIM leverages the strengths of multiple imputation to handle missingness and propagate uncertainty, combined with the power of G-computation to standardize effects across a target population. It is especially useful for estimating marginal effects (e.g., marginal odds ratios) from conditional models, which is crucial for non-collapsible effect measures.

The method integrates two main stages: (1) generation of synthetic datasets (incorporating missing data imputation and counterfactual treatment assignment), and (2) analysis of these synthetic datasets. While it can be implemented with frequentist methods, it naturally accommodates a Bayesian statistical framework, which facilitates principled uncertainty propagation.

\paragraph{Stage 1: Synthetic Data Generation (Imputation and Counterfactual Assignment)}
This stage creates multiple ``complete" synthetic datasets that represent counterfactual scenarios and account for missing data. It typically involves a Bayesian approach for imputation and predictive modeling.

\begin{enumerate}
    \item Missing Data Imputation Model: Fit a multiple imputation model to the observed (potentially incomplete) dataset. This model imputes missing values for $Y$, $A$, and $L$ (or relevant subsets) based on their observed relationships. In a Bayesian context, this involves sampling from the posterior predictive distribution of the missing data given the observed data and specified imputation models.
    Let $D_{obs}$ be the observed data with potentially missing values. We generate $M$ imputed datasets: $D^{(m)}_{\text{complete}}$, for $m=1, \dots, M$.

    \item Outcome Regression Model (Fitted on Imputed Data): For each of the $M$ imputed datasets, fit a parametric outcome regression model relating the outcome $Y$ to treatment $A$ and covariates $L$.
    For a binary outcome, this might be a logistic regression:
    $$ P(Y=1 | A=a, L=l; \theta^{(m)}) = g^{-1}(\alpha^{(m)} + \beta_A^{(m)} a + (\beta_L^{(m)})^T l) $$
    where $g^{-1}(\cdot)$ is the inverse of the link function (e.g., logit inverse). The parameters $(\alpha^{(m)}, \beta_A^{(m)}, \beta_L^{(m)})$ are estimated for each imputed dataset $m$. In a Bayesian setting, these parameters themselves would be drawn from their posterior distribution conditional on $D^{(m)}_{\text{complete}}$.

    \item Counterfactual Data Creation: For each of the $M$ imputed datasets, and for each individual $i$ within that dataset, create two (or more, for multiple regimens) counterfactual copies:
    \begin{itemize}
        \item Set $A=1$: $D_i^{(m, A=1)} = (Y_i^{(m)}, A=1, L_i^{(m)})$
        \item Set $A=0$: $D_i^{(m, A=0)} = (Y_i^{(m)}, A=0, L_i^{(m)})$
    \end{itemize}
    This is conceptual; in practice, you predict outcomes based on these counterfactual assignments.

    \item Prediction of Counterfactual Outcomes: Using the fitted outcome model from step 2 (with parameters $\theta^{(m)}$) and the counterfactual assignments from step 3, predict the outcome for each individual in each counterfactual dataset.
    For each $m=1, \dots, M$ and for each individual $i$:
    $$ \hat{P}(Y=1 | A=1, L_i^{(m)}; \theta^{(m)}) = g^{-1}(\alpha^{(m)} + \beta_A^{(m)} \cdot 1 + (\beta_L^{(m)})^T L_i^{(m)}) $$
    $$ \hat{P}(Y=1 | A=0, L_i^{(m)}; \theta^{(m)}) = g^{-1}(\alpha^{(m)} + \beta_A^{(m)} \cdot 0 + (\beta_L^{(m)})^T L_i^{(m)}) $$
    These predictions are done for each individual within each of the $M$ imputed datasets.
\end{enumerate}

\paragraph{Stage 2: Analysis of Synthetic Datasets and Pooling}

\begin{enumerate}
    \item Marginalization within each Imputed Dataset: For each of the $M$ sets of predictions obtained in Stage 1, calculate the marginal expected outcome by averaging the predicted probabilities across all individuals in that imputed dataset.
    For each $m=1, \dots, M$:
    $$
    \hat{\mathbb{E}}[Y^{A=1}]^{(m)} = \frac{1}{N} \sum_{i=1}^N \hat{P}(Y=1 | A=1, L_i^{(m)}; \theta^{(m)})
    $$
    $$
    \hat{\mathbb{E}}[Y^{A=0}]^{(m)} = \frac{1}{N} \sum_{i=1}^N \hat{P}(Y=1 | A=0, L_i^{(m)}; \theta^{(m)})
    $$

    \item Calculate Effect Measures within each Imputed Dataset: From these marginal means, compute the desired effect measure for each $m$. For example, for a marginal odds ratio:
    $$
    \widehat{\text{OR}}^{(m)} = \frac{\hat{\mathbb{E}}[Y^{A=1}]^{(m)} / (1 - \hat{\mathbb{E}}[Y^{A=1}]^{(m)})}{\hat{\mathbb{E}}[Y^{A=0}]^{(m)} / (1 - \hat{\mathbb{E}}[Y^{A=0}]^{(m)})}
    $$

    \item Pooling Results (Rubin's Rules): The $M$ estimates of the marginal effect measure (e.g., $\widehat{\text{OR}}^{(m)}$) are then combined using Rubin's rules for multiple imputation. This combines the within-imputation variability with the between-imputation variability to provide a single point estimate and a valid measure of total uncertainty (e.g., standard error or credible interval).

    Let $Q$ be the parameter of interest (e.g., $\widehat{\text{OR}}$).
    The pooled point estimate is the average:
    $$ \bar{Q} = \frac{1}{M} \sum_{m=1}^M Q^{(m)} $$
    The total variance $T$ is given by:
    $$ T = \bar{U} + (1 + \frac{1}{M}) B $$
    where $\bar{U} = \frac{1}{M} \sum_{m=1}^M U^{(m)}$ is the within-imputation variance (average of squared standard errors from each imputed analysis), and $B = \frac{1}{M-1} \sum_{m=1}^M (Q^{(m)} - \bar{Q})^2$ is the between-imputation variance.
\end{enumerate}

The ``optimization" in MIM occurs at two levels:
\begin{enumerate}
    \item Imputation Model Fitting: The process of generating multiple imputed datasets involves fitting models (often using MCMC in a Bayesian framework, or chained equations) to estimate the missing data conditional on observed data. This is an optimization to characterize the missing data distribution.
    \item Outcome Model Fitting: For each imputed dataset, the parametric outcome regression model parameters are estimated (e.g., via MLE in a frequentist approach, or MCMC in a Bayesian approach). This is the second layer of optimization, finding the best-fitting parameters for the conditional outcome.
\end{enumerate}
The subsequent steps of prediction, marginalization, and pooling are analytical or simulation-based, not further ``optimization" in the sense of parameter estimation.

Advantages of MIM are Handles Missing Data: Provides a principled way to incorporate uncertainty due to missing data into the causal effect estimation. Estimates Marginal Effects: Correctly estimates marginal causal effects, overcoming the non-collapsibility issue. Propagates Uncertainty: Allows for full propagation of uncertainty from missing data, model parameter estimation, and the marginalization process. Flexibility: Can accommodate various types of outcomes (binary, continuous, time-to-event) and complex covariate structures.


%
\section{Implementation and Software Design}
\label{sec:implementation}

The \pkg{outstandR} package is implemented in \proglang{R} and designed for ease of use while providing powerful statistical capabilities. The core functionality is encapsulated in the main \texttt{outstandR()} function.

\subsection{The outstandR() Function} \label{ssec:outstandr_function}

The primary function \texttt{outstandR()} orchestrates the standardisation process. Its signature is:


\begin{itemize}[noitemsep,topsep=0pt]
    \item \texttt{ipd\_trial}: Individual-level patient data (e.g., for treatments $A$ and $C$), expected in a long format including treatment and outcome columns consistent with the formula object.
    \item \texttt{ald\_trial}: Aggregate-level data (e.g., for treatments $B$ and $C$). This should contain columns for `variable' (covariate name), \texttt{statistic} (e.g., ``mean", ``sd", ``prop", ``sum"), `value' (numerical summary statistic), and `trt' (treatment label, or \texttt{NA} if common covariate distribution is assumed).
    \item \texttt{strategy}: A computation strategy function, which can be one of \texttt{strategy\_maic()}, \texttt{strategy\_stc()}, \texttt{strategy\_gcomp\_ml()}, \texttt{strategy\_gcomp\_stan()}, or \texttt{strategy\_mim()}..
    \item \texttt{ref\_trt}: The reference, common, or anchoring treatment name.
    \item \texttt{CI}: The desired confidence interval level (between 0 and 1, default 0.95).
    \item \texttt{scale}: The relative treatment effect scale. If \texttt{NULL}, it is automatically determined from the model. Options include ``log-odds'', ``log\_relative\_risk'', ``risk\_difference'', ``delta\_z'', ``mean\_difference'', ``rate\_difference'' depending on data type.
    \item \texttt{...}: Additional arguments, currently allowing named arguments to be passed to \texttt{rstanarm::stan\_glm()} via \texttt{strategy\_gcomp\_stan()}.
\end{itemize}
The function returns a list of statistics as an `outstandR' class object, containing means, variances, and confidence intervals for treatment contrasts and absolute values \cite{RemiroAzocar2022c}.

\subsubsection{How the outstandR() function works internally}
The \texttt{outstandR()} function is a high-level wrapper function for carrying out all of the required steps of ITC and population adjustment depending on the chosen method. It is designed to abstract away the task of transporting from the IPD to the ALD population and computing the contrast and absolute statistics of interest. The commonality of code across methods is exploited to simplify the code and adhere to the DRY principle. 

\subsection{The ``strategy'' Family of Functions}
\pkg{outstandR} provides a set of functions called {\it strategies} which define the type of approach to use. The generic signature is

\begin{verbatim}
strategy_*(formula, family, trt_var, ...)
\end{verbatim}

\texttt{formula}: Linear regression formula object, containing prognostic factors (PF) as main effects and effect modifiers (EM) as interactions with the treatment variable e.g \texttt{y $\sim$ pf + trt + trt:em}.

\texttt{family}: A base \proglang{R} \pkg{stats} package `family' object specifying the distribution and link function (e.g., `binomial'). Family objects provide a convenient way to specify the details of the models used by functions such as \texttt{glm()}. See \texttt{stats::family()} for more details.

\texttt{trt\_var}: Treatment variable name; string e.g. \texttt{trt}. If not provided then heuristics are used to guess from the formula.

The asterisk indicates a placeholder for the name of the particular approach from those detailed in the Methods section previously. Table~\ref{tab:strategy_options} gives the strategy functions and optional arguments of their signature.

\begin{table}[htbp] % [htbp] suggests placement: here, top, bottom, page
    \centering
    \caption{Strategy functions and optional arguments available in \pkg{outstandR}}
    \label{tab:strategy_options}
    \begin{tabular}{ll}
        \toprule
        \textbf{Strategy Function} & \textbf{Optional Arguments} \\
        \midrule
        \texttt{strategy\_maic} & \texttt{R} \\
        \addlinespace
        \texttt{strategy\_stc} &  \\
        \addlinespace
        \texttt{strategy\_gcomp\_ml} & \texttt{rho} \\
                            & \texttt{marginal\_distns} \\
                            & \texttt{marginal\_params} \\
                            & \texttt{R} \\
                            & \texttt{N} \\
        \addlinespace
        \texttt{strategy\_gcomp\_stan} & \texttt{rho} \\
                            & \texttt{marginal\_distns} \\
                            & \texttt{marginal\_params} \\
                            & \texttt{N} \\
        \addlinespace
        \texttt{strategy\_mim} & \texttt{rho} \\
                            & \texttt{N} \\
        \bottomrule
    \end{tabular}
\end{table}

The argument definitions are:

\texttt{rho}: A named square matrix of covariate correlations; default \texttt{NA}

\texttt{marginal\_distns}: Marginal distributions names; vector default \texttt{NA}

\texttt{marginal\_params}: Marginal distributions parameters; list of lists, default \texttt{NA}

\texttt{R}: Number of resamples used in non-parametric bootstrap; default 1000

\texttt{N}: Synthetic sample size for G-computation; default

\subsection{Outcome scales}
If we do not explicitly specify the outcome scale, the default is that
used to fit to the data in the regression model. As we saw, in this
case, the default is log-odds ratio corresponding to the ``logit'' link
function for binary data. However, we can change this to some other
scale which may be more appropriate for a particular analysis. So far
implemented in the package, the links and their corresponding relative
treatment effect scales are as follows:

\begin{table}[h!]
    \centering
    \caption{Data types and associated scales available in \pkg{outstandR}.}
    \label{tab:scales}
    \begin{tabular}{llll}
    \hline
    Data Type & Model & Scale & Argument \\
    \hline
    Binary & \texttt{logit} & Log-odds ratio & \texttt{log\_odds} \\
    Count & \texttt{log} & Log-risk ratio & \texttt{log\_relative\_risk} \\
    Continuous & \texttt{mean} & Mean difference & \texttt{risk\_difference} \\
    \hline
    \end{tabular}
\end{table}

The full list of possible transformed treatment effect scales will be:
log-odds ratio, log-risk ratio, mean difference, risk difference, hazard ratio, hazard difference.

\subsection{Related R packages}
Related \proglang{R} packages we are aware of consist of more general-purpose tools not specific to ITC, and that implement single, specific methods rather than a suite of methods with a common interface. The \pkg{marginaleffects} \citep{aari_marginaleffects_2024} package is not designed for population adjustment between studies. \pkg{stdReg2} focuses on standardising outcomes with a single data set \citep{g_sofer_2023_10022204}.
For G-formula implementations, \pkg{gfoRmula} can estimate effects in the presence of
time-varying treatments and confounders \citep{sjolander2023gformula}. It is designed to estimate causal effects from longitudinal data with one study. \pkg{gFormulaMI} employs multiple imputation using the \pkg{mice} package \citep{Sterne2023}. Finally, \pkg{maicplus} is a specialist ITC package but focused only on the MAIC approach \citep{maicplus_2024}.

\subsection{Software architecture and S3 class design}
We adopted a functional object-oriented design using \proglang{R}'s S3 system to balance modularity with ease of use. The core design philosophy separates the \emph{definition} of the statistical approach from its \emph{execution}.

This is achieved through ``strategy'' objects. Functions such as \code{strategy_maic()}, \code{strategy_stc()}, and \code{strategy_gcomp_ml()} do not perform any immediate calculation. Instead, they validate the user's hyperparameters (e.g., formula, error distribution, bootstrap iterations) and return a structured list object with a specific class attribute (e.g., \code{c("maic", "strategy", "list")}).

This design offers two key advantages:
\begin{itemize}
  \item \textbf{Unified Interface:} The main wrapper function \code{outstandR()} remains agnostic to the specific method being used. It accepts a \code{strategy} object and delegates the computational heavy lifting to the appropriate backend. This simplifies the user API, as the function signature for \code{outstandR()} remains consistent regardless of whether the user is performing a simple MAIC or a complex Bayesian G-computation.
  \item \textbf{Extensibility:} New methods can be added by simply defining a new constructor function (e.g., \code{strategy_newmethod()}) and a corresponding S3 method for the internal generic \code{calc_IPD_stats()}. This allows the package to grow without modifying the core orchestration logic.
\end{itemize}

\subsection{Internal method dispatch}
The package relies on standard S3 method dispatch to route the analysis to the correct computational engine. The internal generic function \code{calc_IPD_stats()} is the central dispatch point.

When \code{outstandR()} is called, it pre-processes the input data and then calls \code{calc_IPD_stats(strategy, analysis_params)}. \proglang{R}'s S3 system inspects the class of the \code{strategy} object and dispatches to the specific implementation:

\begin{itemize}
  \item \code{calc_IPD_stats.maic}: Handles propensity score weighting and bootstrap estimation.
  \item \code{calc_IPD_stats.stc}: Manages outcome regression modeling and prediction.
  \item \code{calc_IPD_stats.gcomp_ml}: Orchestrates the simulation of pseudo-populations and maximum likelihood estimation.
  \item \code{calc_IPD_stats.gcomp_stan}: Interfaces with the \proglang{Stan} backend for Bayesian inference.
\end{itemize}

To adhere to the DRY (Don't Repeat Yourself) principle, we utilized a functional programming pattern with the \code{IPD_stat_factory()} helper. Since many strategies require similar post-processing---such as calculating the Average Treatment Effect (ATE) from arm-specific means and formatting the output list---\code{IPD_stat_factory} wraps the specific calculation functions. It standardizes the output structure (means, variances, confidence intervals) across all methods, ensuring that the \code{outstandR} return object is consistent and predictable for the user.

\subsection{Computational efficiency and memory management in G-computation}
G-computation involves simulating large pseudo-populations to numerically approximate the integration over the covariate space. This process can be computationally intensive and memory-demanding, particularly when combined with bootstrapping (in the maximum likelihood approach) or MCMC sampling (in the Bayesian approach).

In \pkg{outstandR}, the memory footprint is largely driven by the size of the synthetic cohort used to represent the target population. To manage this, we expose the simulation size parameter \code{N} directly in the strategy functions (\code{strategy_gcomp_ml} and \code{strategy_gcomp_stan}).

\begin{itemize}
  \item \textbf{User-Controlled Precision vs.\ Performance:} Users can explicitly tune \code{N} (defaulting to 1,000) to balance the precision of the marginal effect estimate against available memory resources.
  \item \textbf{Simulation Logic:} The internal function \code{simulate_ALD_pseudo_pop()} generates the synthetic covariate matrix $X^*$ using a multivariate normal copula. To calculate marginal effects, the package generates counterfactual datasets by duplicating $X^*$ for each treatment arm. While \proglang{R}'s copy-on-modify semantics help minimize overhead during the initial generation, the creation of counterfactuals (setting treatment columns to 0 and 1) forces memory allocation. By allowing users to control \code{N}, \pkg{outstandR} enables execution on standard laptops for moderate precision while supporting high-precision runs on high-memory servers.
\end{itemize}

\section{Using the outstandR package}
\label{sec:examples}
This section gives a practical demonstration of the core functionality available in the package.

\subsection{Available internal data sets}
To illustrate the use of \pkg{outstandR}, we demonstrate its application with synthetic datasets.
These are contained in the \pkg{outstandR} package and can be loaded via the \texttt{data()} function.
The naming convention that we use is \newline \\
\texttt{<comparator><reference>\_<level>\_<outcome>\_<covariates>} \newline
\\
where comparator and reference are the treatment arm names, level is either IPD for individual-level patient data or ALD for aggregate-level data, outcome is binary or continuous and covariates is binary, continuous or mixed.
For example, \texttt{AC\_IPD\_binY\_binX} is the name of the data set for treatments $A$ vs $C$, patient-level data, and both binary-valued outcome and covariates.

\subsection{Installation}
First, install the development version from GitHub using R-universe:
\begin{verbatim}
install.packages("outstandR",
                 repos = c("https://statisticshealtheconomics.r-universe.dev",
                 "https://cloud.r-project.org"))
\end{verbatim}


\subsection{Illustrative Examples}
Attach the package to the current environment.

\begin{verbatim}
library(outstandR)
\end{verbatim}

\subsubsection{Binary Outcome Data, Binary Covariates}
Load the aggregate level and individual-patient data in to the environment.

\begin{verbatim}
data(AC_IPD_binY_binX) # AC patient-level data
data(BC_ALD_binY_binX) # BC aggregate-level data
\end{verbatim}

Inspecting the IPD, we can see that each row corresponds to an individual and it has four covariates, two effect modifiers only and two prognostic factors only. These are all continuous valued and the outcome is binary values 0, 1.

\begin{verbatim}
# individual-level data
head(AC_IPD_binY_contX)

##   id PF_cont_1   PF_cont_2   EM_cont_1   EM_cont_2 trt y    true_eta
## 1  1 0.0826453  0.71645068  0.66422255 -0.09202652   A 0 -1.58891404
## 2  2 0.6957433  0.38608723  0.79875530  0.74880584   C 1  0.14986776
## 3  3 0.6516889  0.50572444  0.39944706 -0.07230380   A 0 -1.43868552
## 4  4 0.2165975  0.35783123  0.25957984 -0.55405590   A 0 -2.09172428
## 5  5 0.2605580 -0.05826527 -0.10696235  0.33762456   A 0 -2.13936317
## 6  6 0.1110818  0.64041589  0.04867791  0.34980415   C 1 -0.07910151
\end{verbatim}

Similarly, the ALD has the corresponding covariates but not as separate columns (wide format) but in a long format with all covariate names in the \texttt{variable} column.  Notice that the treatment column \texttt{trt} is NA for the covariate because we assume that the covariate distribution is the same between treatment arms.

\begin{verbatim}
# aggregate-level data
BC_ALD_binY_contX

## # A tibble: 16 × 4
##    variable  statistic   value trt  
##    <chr>     <chr>       <dbl> <chr>
##  1 EM_cont_1 mean        0.651 <NA> 
##  2 EM_cont_1 sd          0.391 <NA> 
##  3 EM_cont_2 mean        0.592 <NA> 
##  4 EM_cont_2 sd          0.416 <NA> 
##  5 PF_cont_1 mean        0.653 <NA> 
##  6 PF_cont_1 sd          0.371 <NA> 
##  7 PF_cont_2 mean        0.583 <NA> 
##  8 PF_cont_2 sd          0.437 <NA> 
##  9 y         mean        0.615 C    
## 10 y         sd          0.490 C    
## 11 y         sum        40     C    
## 12 y         mean        0.274 B    
## 13 y         sd          0.448 B    
## 14 y         sum        37     B    
## 15 <NA>      N          65     C    
## 16 <NA>      N         135     B
\end{verbatim}

The next thing is to define the formula which specifies the relationship between the covariates, treatments and outcome. This is independent of the particular type of data and so we don't include an link function at this stage.

For this example our formula is given by

\begin{verbatim}
# linear formula for the outcome model
lin_form <- as.formula(
   "y ~ PF_cont_1 + PF_cont_2 + trt + trt:EM_cont_1 + trt:EM_cont_2")
\end{verbatim}
That is, the main effects are the prognostic factors (\texttt{PF\_cont\_1} and \texttt{PF\_cont\_2}) and the treatment  \texttt{trt}. The interaction terms are with the treatment defined as effect modifiers (\texttt{EM\_cont\_1} and \texttt{EM\_cont\_1}). 

\paragraph{Running oustandR()}
Having completed the set-up steps we are ready to proceed with performing the indirect treatment comparison with \texttt{outstandR()}. For these data we shall use \texttt{family = binomial(link = "logit")} for the logistic model.
First, for an MAIC analysis use \texttt{strategy\_maic()}.

\begin{verbatim}
outstandR_maic <-
  outstandR(AC_IPD_binY_contX, BC_ALD_binY_contX,
            strategy = strategy_maic(
              formula = lin_form,
              family = binomial(link = "logit")))
\end{verbatim}

Looking at the results returns a list object.

\begin{verbatim}
str(outstandR_maic)

## List of 2
##  $ contrasts:List of 3
##   ..$ means    :List of 3
##   .. ..$ AB: num 0.62
##   .. ..$ AC: num -0.824
##   .. ..$ BC: num -1.44
##   ..$ variances:List of 3
##   .. ..$ AB: num 0.239
##   .. ..$ AC: num 0.137
##   .. ..$ BC: num 0.102
##   ..$ CI       :List of 3
##   .. ..$ AB: num [1:2] -0.338 1.579
##   .. ..$ AC: num [1:2] -1.5491 -0.0985
##   .. ..$ BC: num [1:2] -2.071 -0.817
##  $ absolute :List of 2
##   ..$ means    :List of 2
##   .. ..$ A: Named num 0.245
##   .. .. ..- attr(*, "names")= chr "mean_A"
##   .. ..$ C: Named num 0.422
##   .. .. ..- attr(*, "names")= chr "mean_C"
##   ..$ variances:List of 2
##   .. ..$ A: Named num 0.00212
##   .. .. ..- attr(*, "names")= chr "mean_A"
##   .. ..$ C: Named num 0.00527
##   .. .. ..- attr(*, "names")= chr "mean_C"
##  - attr(*, "CI")= num 0.95
##  - attr(*, "ref_trt")= chr "C"
##  - attr(*, "scale")= chr "log_odds"
##  - attr(*, "model")= chr "binomial"
##  - attr(*, "class")= chr [1:2] "outstandR" "list"
\end{verbatim}

Results are presented as contrasts and absolute values. For each of these the means, variances and confidence intervals are provided.
Additional attributes (\texttt{attr}) are included for the confidence interval range, the reference treatment name, the outcome scale and the type of likelihood model.

To help with interpreting the results, \pkg{outstandR} includes a convenience \texttt{print} method.

\begin{verbatim}
print(outstandR_maic)

## Object of class 'outstandR' 
## Model: binomial 
## Scale: log_odds 
## Common treatment: C 
## Individual patient data study: AC 
## Aggregate level data study: BC 
## Confidence interval level: 0.95 
## 
## Contrasts:
## 
## # A tibble: 3 × 5
##   Treatments Estimate Std.Error lower.0.95 upper.0.95
##   <chr>         <dbl>     <dbl>      <dbl>      <dbl>
## 1 AB            0.620     0.239     -0.338     1.58  
## 2 AC           -0.824     0.137     -1.55     -0.0985
## 3 BC           -1.44      0.102     -2.07     -0.817 
## 
## Absolute:
## 
## # A tibble: 2 × 5
##   Treatments Estimate Std.Error lower.0.95 upper.0.95
##   <chr>         <dbl>     <dbl> <lgl>      <lgl>     
## 1 A             0.245   0.00212 NA         NA        
## 2 C             0.422   0.00527 NA         NA
\end{verbatim}

This contains the same information as the raw output but in a more presentable format. The first block of text provides modelling details, and the following tibbles show the contrast and absolute effects, respectively.

Similarly, the STC, G-computation and MIM analyses are

\begin{verbatim}
outstandR_stc <-
  outstandR(AC_IPD_binY_contX, BC_ALD_binY_contX,
            strategy = strategy_stc(
              formula = lin_form,
              family = binomial(link = "logit")))

outstandR_gcomp_ml <-
  outstandR(AC_IPD_binY_contX, BC_ALD_binY_contX,
            strategy = strategy_gcomp_ml(
              formula = lin_form,
              family = binomial(link = "logit")))

outstandR_gcomp_stan <-
  outstandR(AC_IPD_binY_contX, BC_ALD_binY_contX,
            refresh = 0,  # turns off the iteration updates
            strategy = strategy_gcomp_stan(
              formula = lin_form,
              family = binomial(link = "logit")))

outstandR_mim <-
  outstandR(AC_IPD_binY_contX, BC_ALD_binY_contX,
            refresh = 0,
            strategy = strategy_mim(
              formula = lin_form,
              family = binomial(link = "logit")))
\end{verbatim}

\subsubsection{Continuous Outcome Data, Mixture of Binary and Continuous Covariates}

Analogous to the first example with binary outcomes and continuous covariates, we now extend this for the case of continuous outcomes and mixed covariates. The exact same steps are followed.

First read in the data.

\begin{verbatim}
# individual-level data
head(AC_IPD_contY_mixedX)

##   id        X1 X2        X3 X4 trt          y   true_eta
## 1  1 0.3066158  0 1.0794085  0   C -2.3321553 -0.9789591
## 2  2 0.5961703  1 0.6164710  0   A -0.7473152 -0.8840576
## 3  3 0.1496635  1 0.5737457  0   A -1.2441749 -1.2402780
## 4  4 0.1393548  0 0.4687338  0   A -0.7905105 -0.1785051
## 5  5 0.5672248  0 0.5305725  0   A -3.3441703 -1.3985394
## 6  6 0.4897103  1 0.1289250  1   A -1.2286456 -0.3974177

# aggregate-level data
BC_ALD_contY_mixedX

## # A tibble: 16 × 4
##    variable statistic   value trt  
##    <chr>    <chr>       <dbl> <chr>
##  1 X1       mean        0.608 <NA> 
##  2 X1       sd          0.436 <NA> 
##  3 X2       prop        0.72  <NA> 
##  5 X3       mean        0.598 <NA> 
##  6 X3       sd          0.385 <NA> 
##  7 X4       prop        0.51  <NA> 
##  9 y        mean        0.493 C    
## 10 y        sd          1.08  C    
## 11 y        sum        36.0   C    
## 12 y        mean       -0.434 B    
## 13 y        sd          1.39  B    
## 14 y        sum       -55.1   B    
## 15 <NA>     N          73     C    
## 16 <NA>     N         127     B
\end{verbatim}

The formats are the same as the previous example but now in the ALD the binary-valued covariates have \texttt{statistic} entry \texttt{prop}. \texttt{X1} and \texttt{X3} are continuous and \texttt{X2} and \texttt{X4} are binary.

The formula for this example is also a little more complicated. We shall assume that some of the covariate can be both prognostic factors and effect modifiers. 

\begin{verbatim}
lin_form <- as.formula("y ~ X1 + X2 + X3 + trt + trt:(X1 + X2 + X4)")
\end{verbatim}

A difference from the first example is the \texttt{family = gaussian(link = "identity")} argument indicating a Normal linear regression.

\begin{verbatim}
outstandR_maic <-
  outstandR(AC_IPD_contY_mixedX, BC_ALD_contY_mixedX,
            strategy = strategy_maic(
              formula = lin_form,
              family = gaussian(link = "identity")))
\end{verbatim}

Again, if we interrogate the output we see the same list format. Differences to the first example are the model and scale attributes correspond to the input data used. That is, the default output contrast is the mean difference.

\begin{verbatim}
str(outstandR_maic)

## List of 2
##  $ contrasts:List of 3
##   ..$ means    :List of 3
##   .. ..$ AB: num 0.58
##   .. ..$ AC: num -0.346
##   .. ..$ BC: num -0.926
##   ..$ variances:List of 3
##   .. ..$ AB: num 0.14
##   .. ..$ AC: num 0.109
##   .. ..$ BC: num 0.0311
##   ..$ CI       :List of 3
##   .. ..$ AB: num [1:2] -0.154 1.314
##   .. ..$ AC: num [1:2] -0.994 0.301
##   .. ..$ BC: num [1:2] -1.272 -0.581
##  $ absolute :List of 2
##   ..$ means    :List of 2
##   .. ..$ A: Named num -0.336
##   .. .. ..- attr(*, "names")= chr "mean_A"
##   .. ..$ C: Named num 0.0104
##   .. .. ..- attr(*, "names")= chr "mean_C"
##   ..$ variances:List of 2
##   .. ..$ A: Named num 0.0381
##   .. .. ..- attr(*, "names")= chr "mean_A"
##   .. ..$ C: Named num 0.0594
##   .. .. ..- attr(*, "names")= chr "mean_C"
##  - attr(*, "CI")= num 0.95
##  - attr(*, "ref_trt")= chr "C"
##  - attr(*, "scale")= chr "mean_difference"
##  - attr(*, "model")= chr "gaussian"
##  - attr(*, "class")= chr [1:2] "outstandR" "list"

print(outstandR_maic)

## Object of class 'outstandR' 
## Model: gaussian 
## Scale: mean_difference 
## Common treatment: C 
## Individual patient data study: AC 
## Aggregate level data study: BC 
## Confidence interval level: 0.95 
## 
## Contrasts:
## 
## # A tibble: 3 × 5
##   Treatments Estimate Std.Error lower.0.95 upper.0.95
##   <chr>         <dbl>     <dbl>      <dbl>      <dbl>
## 1 AB            0.580    0.140      -0.154      1.31 
## 2 AC           -0.346    0.109      -0.994      0.301
## 3 BC           -0.926    0.0311     -1.27      -0.581
## 
## Absolute:
## 
## # A tibble: 2 × 5
##   Treatments Estimate Std.Error lower.0.95 upper.0.95
##   <chr>         <dbl>     <dbl> <lgl>      <lgl>     
## 1 A           -0.336     0.0381 NA         NA        
## 2 C            0.0104    0.0594 NA         NA
\end{verbatim}

The corresponding calls to \texttt{outstandR()} for the other analyses follow similarly.

\begin{verbatim}
outstandR_stc <-
  outstandR(AC_IPD_contY_mixedX, BC_ALD_contY_mixedX,
            strategy = strategy_stc(
              formula = lin_form,
              family = gaussian(link = "identity")))

outstandR_gcomp_ml <-
  outstandR(AC_IPD_contY_mixedX, BC_ALD_contY_mixedX,
            strategy = strategy_gcomp_ml(
              formula = lin_form,
              family = gaussian(link = "identity")))

outstandR_gcomp_stan <-
  outstandR(AC_IPD_contY_mixedX, BC_ALD_contY_mixedX,
            refresh = 0,  # turns off the iteration updates
            strategy = strategy_gcomp_stan(
              formula = lin_form,
              family = gaussian(link = "identity")))

outstandR_mim <-
  outstandR(AC_IPD_contY_mixedX, BC_ALD_contY_mixedX,
            refresh = 0,
            strategy = strategy_mim(
              formula = lin_form,
              family = gaussian(link = "identity")))
\end{verbatim}

\subsection{User-Specified Outcome Scales}
By default, \pkg{outstandR} reports treatment effects on the scale corresponding to the regression model's link function. For instance, a logistic regression specified with \texttt{family = binomial(link = "logit")} naturally yields estimates in log-odds ratios, while a linear regression yields mean differences. However, health economic models often require parameters on specific scales, such as Risk Differences (RD) or Relative Risks (RR), to facilitate interpretation or align with decision-analytic model inputs. The \texttt{scale} argument in \texttt{outstandR()} allows users to request these alternative summary statistics directly. When converting between relative measures (e.g., Odds Ratios) and absolute measures (e.g., Risk Differences), the package automatically estimates the necessary baseline risk ($P_0$) from the reference treatment arm in the IPD study. For example, to obtain a Risk Difference instead of the default Log-Odds Ratio for the binary outcome analysis (previously shown in Section~\ref{sec:examples}), we specify \texttt{scale = "risk\_difference"}:

\begin{verbatim}
# Requesting Risk Difference instead of Log-Odds Ratio
outstandR_maic_rd <- outstandR(
  ipd_trial = AC_IPD_binY_contX, 
  ald_trial = BC_ALD_binY_contX,
  strategy = strategy_maic(
    formula = lin_form,
    family = binomial(link = "logit")
  ),
  scale = "risk_difference"
)

print(outstandR_maic_rd)
\end{verbatim}

The output now displays the contrast estimates on the probability scale:

\begin{verbatim}
## Object of class 'outstandR'
## Model: binomial
## Scale: risk_difference
## Contrasts:
## # A tibble: 3 x 5
##   Treatments Estimate Std. Error lower.0.95 upper.0.95
##   <chr>         <dbl>      <dbl>      <dbl>      <dbl>
## 1 AB           0.151     0.052     0.049      0.253
## 2 AC          -0.198     0.031    -0.259     -0.137
\end{verbatim}

Supported scales include ``log\_odds", ``odds\_ratio", ``log\_relative\_risk", ``relative\_risk", ``risk\_difference", ``mean\_difference", and ``rate\_difference". The package determines the validity of a conversion path based on the underlying data distribution; for instance, requesting a "mean\_difference" for a binary outcome model will raise a validation error.

\section{Conclusion}
\label{sec:conclusion}

The \pkg{outstandR} package offers a comprehensive and flexible toolset for addressing indirect treatment comparisons, especially in the challenging context of limited individual patient data. Its key strengths include:
\begin{itemize}
    \item Versatility: Support for multiple standardisation methods (MAIC, STC, G-computation, MIM) allows users to choose the most appropriate approach for their data and assumptions.
    \item Uncertainty Quantification: The Bayesian G-computation approach provides a robust framework for propagating and quantifying uncertainty.
    \item Practical Relevance: Directly addresses a critical need in health economics and evidence synthesis for comparing treatments when direct trial data is unavailable.
\end{itemize}
While powerful, the package relies on the correct specification of outcome models and the availability of sufficient aggregate-level data for reliable estimation. Future work will focus on expanding the range of supported outcome models and exploring more advanced covariate simulation techniques.


\bibliographystyle{plainnat} % Or another suitable bibliography style like jss
\bibliography{references} % Assumes .bib file is named 'references.bib'

\end{document}